Positivity-Preserving Adaptive Runge--Kutta Methods

### 2021-01-31 nusslein_vsrp/latex/response_to_reviewers.tex

\begin{enumerate}[label=\arabic*.]
  \setcounter{enumi}{1}
  \item Bound-preserving adaptive Runge–Kutta methods
  \begin{enumerate}[label=\alph*)]
    \item \emph{%
    In equation (2.14b), it is weird to read $-0.01 u_2$ and $- 0.05 u_2$
    in the same line. Is it an error? The same thing happens in equations (2.14c)
    with $u_3$, in (5.5b) with $u_2$, and in (5.5c) with $u_3$.
    }

    We added an explanation why we use this seemingly weird notation of the
    equations [lines~{line:prod-dest}ff.].

@@

    \item \emph{%
    In Figure 1, in the difference $\| \tilde b - b \|$, a tilde is missing.
    }

    \revA{TODO} %TODO
  \end{enumerate}

  \item Selection of the modified weights
  \begin{enumerate}[label=\alph*)]
    \item \emph{%
    In section 3.2, two different approaches to adapt the weights are considered:
    Free adaptation and Convex adaptation. At the same time, it is explained that
    ``...we can additionally impose either or both of the following strategies \dots''.
    I think that the first strategy is nothing but the second approach (Convex
    adaptation). I suggest making this part clearer in order to avoid confusion
    between approaches and strategies.
    }

    We modified this part to be more clear [lines~\ref{line:strategies}ff.].
  \end{enumerate}

@@

  \item Numerical experiments
  \begin{enumerate}[label=\alph*)]
    \item \emph{%
    In the third line of Section 5.1, RHS appears for the first time without
    explaining its meaning.
    }

    We changed this sentence and introduced the abbreviation RHS
    [line~\ref{line:RHS}].

    \item \emph{%
    In section 5.1, just before the formula (5.1), it is said ``Since the
    positivity of the stage values is not ensured, the right hand side has to be
    defined for negative values. If this is not the case, the right hand side
    has to to be adapted accordingly''. This is a bit confusing, particularly
    the text ``...has to be defined for negative values...''. I suggest that
    you make this part clearer.
    }

    We tried to clarify this part by giving an example and adding further
    explanations [lines~\ref{line:RHS-negative-values}ff.].

### 2021-02-10 nusslein_vsrp/latex/response_to_reviewers.tex

  \item Selection of the modified weights
  \begin{enumerate}[label=\alph*)]
    \item \emph{%
    In section 3.2, two different approaches to adapt the weights are considered:
    Free adaptation and Convex adaptation. At the same time, it is explained that
    ``...we can additionally impose either or both of the following strategies \dots''.
    I think that the first strategy is nothing but the second approach (Convex
    adaptation). I suggest making this part clearer in order to avoid confusion
    between approaches and strategies.
    }

    We modified this part to be more clear [lines~\ref{line:strategies}ff.].
  \end{enumerate}

  \item Properties of adaptive RKMs
  \begin{enumerate}[label=\alph*)]
    \item \emph{%
    Figure 3a, with free adaptation of the weights $\tilde b$, does not give enough
    information about the stability regions of the adapted methods. In my opinion,
    it would be clearer if each stability curve was labeled with the value of
    the difference $\| \tilde b - b \|$, accordingly with the values in Figure 4b.
    In this way we could see the progress of the stability regions during the
    numerical integration process.
    }

    \revA{Review the following:} %TODO
    We found an additional issue: The step size used for this plot was not $\Delta t = 0.015$ as for Figure 4a. We corrected the step size. Therefore the shape of the plotted stability regions changed. 
	To indicate the value of $\| \tilde b - b \|$ for the different stability regions the boundaries were colored using a colormap.  
	We removed the filling pattern to improve clarity.

### 2021-02-28 nusslein_vsrp/latex/response_to_reviewers.tex

\begin{enumerate}[label=\arabic*.]
  \setcounter{enumi}{1}
  \item Bound-preserving adaptive Runge–Kutta methods
  \begin{enumerate}[label=\alph*)]
    \item \emph{%
    In equation (2.14b), it is weird to read $-0.01 u_2$ and $- 0.05 u_2$
    in the same line. Is it an error? The same thing happens in equations (2.14c)
    with $u_3$, in (5.5b) with $u_2$, and in (5.5c) with $u_3$.
    }

    We added an explanation why we use this seemingly weird notation of the
    equations [lines~\ref{line:prod-dest}ff.].

@@

  \item Selection of the modified weights
  \begin{enumerate}[label=\alph*)]
    \item \emph{%
    In section 3.2, two different approaches to adapt the weights are considered:
    Free adaptation and Convex adaptation. At the same time, it is explained that
    ``...we can additionally impose either or both of the following strategies \dots''.
    I think that the first strategy is nothing but the second approach (Convex
    adaptation). I suggest making this part clearer in order to avoid confusion
    between approaches and strategies.
    }

    We modified this part to be more clear [lines~\ref{line:strategies}ff.].
  \end{enumerate}

  \item Properties of adaptive RKMs
  \begin{enumerate}[label=\alph*)]
    \item \emph{%
    Figure 3a, with free adaptation of the weights $\tilde b$, does not give enough
    information about the stability regions of the adapted methods. In my opinion,
    it would be clearer if each stability curve was labeled with the value of
    the difference $\| \tilde b - b \|$, accordingly with the values in Figure 4b.
    In this way we could see the progress of the stability regions during the
    numerical integration process.
    }

    \revA{TODO: Review the following:} %TODO
    We found an additional issue: The step size used for this plot was not $\Delta t = 0.015$ as for Figure 4a. We corrected the step size. Therefore the shape of the plotted stability regions changed. 
	To indicate the value of $\| \tilde b - b \|$ for the different stability regions the boundaries were colored using a colormap.  
	We removed the filling pattern to improve clarity.

@@

    \item \emph{%
    After formula (5.2) ``...In Figure 4a the results for $\Delta t = 0.015$
    are plotted for different time steps...'' should be ``...In Figure 4a the
    results for $\Delta t = 0.015$ are plotted for different values of time $t$...''
    }

    We fixed this sentence [line~\ref{line:time-vs-dt}].

    \item \emph{%
    Figure 5, about the non-stiff problem with fixed stepsize, is quite interesting
    and gives information about the convergence of the adapted mehtod. For
    $\Delta t = 0.0082$ the unaltered method leads to positive solutions. For
    $\Delta t = 0.015$ the original method leads to negative values in the
    interval $[0, 0.375]$ and the weights are altered by solving an LP problem.
    Is this extra time consuming a significant cost or could it be an economical
    alternative to using excessively small step sizes? A further explanation
    about this issue should be added.
    }

    \revA{TODO: Stephan wants to run some numerical experiments} %TODO

### 2021-03-01 nusslein_vsrp/latex/overview.tex
fix missing values
Added some comment on the cost of solving the LP for the non stiff case

A test problem similar to \cite{shampine_non-negative_2005} is the PDE
\begin{equation}
\label{eq:advection-decay}
\begin{aligned}
  \frac{\partial u(t,x)}{\partial t}
  &=
  -a \frac{\partial u(t,x)}{\partial x} - K u(t,x),
  && x \in (0, 1), t \in (0,1),
  \\
  u(t,0) &= 1,
  && t \in (0,1],
  \\
  u(0,x) &= 0,
  && x \in [0,1],
\end{aligned}
\end{equation}
which consists of an advection part and an exponential decay.
The numerical approximation uses the method of lines and a first order
upwind finite difference semidiscretization with $N = 100$ points.
This leads to the positivity preserving ODE
\begin{equation}
\label{eq:advection-decay-semidiscrete}
\frac{\mathrm d}{\mathrm d t} u_i = \frac{a}{\Delta x} \left( u_{i-1} - u_i \right) - K \revA{u_{i}}.
\end{equation}
The parameters are set to $a=1$ and $K=1$.
We use the Dormand--Prince RK5 method and adapt the weights using the free adaptation.
In Figure\,\ref{fig:sol_AdDe} the results for $\Delta t = 0.015$ are plotted for different \linelabel{line:time-vs-dt}\revA{of time $t$}. We can see that the solution approaches an exponential function with $t \rightarrow \infty$.
In Figure\,\ref{fig:weights_AdDe} the used weights are plotted. For $t\leq 0.375$ the weights are altered. For $t > 0.375$ the original weights lead to a positive solution.
\begin{figure}
\centering
\begin{subfigure}[t]{0.45\textwidth}
\centering
\includegraphics[width=1\textwidth]{plots/Advection_Decay.pdf}
\caption{Numerical solution at different times.}
\label{fig:sol_AdDe}
\end{subfigure}%
~
\begin{subfigure}[t]{0.45\textwidth}
\centering
\includegraphics[width=1\textwidth]{plots/b_Advection_Decay.pdf}\\
\caption{Adaptation of the weights.}
\label{fig:weights_AdDe}
\end{subfigure}
\caption{Numerical results for the advection decay problem \eqref{eq:advection-decay}.}
\end{figure}

@@

The unaltered method has the order $p=5$.
The values marked with a cross denote the numerical experiments that required an adaption of the weights.
Even though the order is reduced, most errors are still close to the error of the unaltered method.

>>>>

A test problem similar to \cite{shampine_non-negative_2005} is the PDE
\begin{equation}
\label{eq:advection-decay}
\begin{aligned}
  \frac{\partial u(t,x)}{\partial t}
  &=
  -a \frac{\partial u(t,x)}{\partial x} - K u(t,x),
  && x \in (0, 1), t \in (0,1),
  \\
  u(t,0) &= 1,
  && t \in (0,1],
  \\
  u(0,x) &= 0,
  && x \in [0,1],
\end{aligned}
\end{equation}
which consists of an advection part and an exponential decay.
The numerical approximation uses the method of lines and a first order
upwind finite difference semidiscretization with $N = 100$ points.
This leads to the positivity preserving ODE
\begin{equation}
\label{eq:advection-decay-semidiscrete}
\frac{\mathrm d}{\mathrm d t} u_i = \frac{a}{\Delta x} \left( u_{i-1} - u_i \right) - K \revA{u_{i}}.
\end{equation}
The parameters are set to $a=1$ and $K=1$.
We use the Dormand--Prince RK5 method and adapt the weights using the free adaptation.
In Figure\,\ref{fig:sol_AdDe} the results for $\Delta t = 0.015$ are plotted for different \linelabel{line:time-vs-dt}\revA{values of time $t$}. We can see that the solution approaches an exponential function with $t \rightarrow \infty$.
In Figure\,\ref{fig:weights_AdDe} the used weights are plotted. For $t\leq 0.375$ the weights are altered. For $t > 0.375$ the original weights lead to a positive solution.
\begin{figure}
\centering
\begin{subfigure}[t]{0.45\textwidth}
\centering
\includegraphics[width=1\textwidth]{plots/Advection_Decay.pdf}
\caption{Numerical solution at different times.}
\label{fig:sol_AdDe}
\end{subfigure}%
~
\begin{subfigure}[t]{0.45\textwidth}
\centering
\includegraphics[width=1\textwidth]{plots/b_Advection_Decay.pdf}\\
\caption{Adaptation of the weights.}
\label{fig:weights_AdDe}
\end{subfigure}
\caption{Numerical results for the advection decay problem \eqref{eq:advection-decay}.}
\end{figure}

@@

The unaltered method has the order $p=5$.
The values marked with a cross denote the numerical experiments that required an adaption of the weights.
Even though the order is reduced, most errors are still close to the error of the unaltered method.
\linelabel{line:cost}\revA{
In this example the cost of computing the stage values is small. Additionally a small reduction of $\dt$ is enough to obtain positive results.  
If we compare the computation time we can see that using a smaller $\dt$ is faster than the adaption. 
Therefore solving the LP is no economical alternative to reducing the step size.}
